Fail Faster

### 2025-03-21 paper/paper.tex
put back

This is due to two major sources of inefficiency.
First, the same high-level design approach that gives generator libraries their
flexibility also introduces layered abstractions and indirections that are difficult for
compilers to optimize, shifting work to run time.
Second, the \rand that a generator uses to make choices can
constitute an unexpectedly high proportion of its runtime. Combined, these factors limit 
the effectiveness of PBT.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi

We lay a path forward.
First, we show how to use \textit{multi-stage programming}---a classic technique for 
building fast DSLs “without regret”~\cn{}---to eliminate the overhead of 
many common generator abstractions.
Second, we characterize the impact of \rand choice through a natural experiment 
that isolates the performance contribution of the \rand within a generator, underscoring its role as a first-class design consideration.
Taken together, these insights can be applied to create tomorrow's generator libraries: those that 
combine high-level design with low-level performance.

To demonstrate the effectiveness and generality of our approach\bcp{It
feels a bit grand to call what we've done an ``approach''---it's
really just a single performance-enhancing technique, staging, plus
the observation that the RNG matters and you should use the fastest
one you can.}, which we call \name, we implement staged PBT libraries in multiple languages.
In OCaml, we build \camlname{}, based on \bq{}---a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
In Scala 3, we build \scalaname{}, based on ScalaCheck---a very
mature PBT library and the {\em de facto} choice\bcp{the wording here
  implies that there are other choices in Scala, some of which might
  be reasonable, and tht base\_quickcheck is not the de facto choice
  in OCaml...} in Scala \cn{}.
In both languages, our libraries are direct drop-in
replacements\bcp{that's not quite accurate, is it? I mean, they have
  the same functionality, but can you take a generator that works with
base\_quickcheck and just link it against our version without changing
the generator itself? That's what ``drop-in replacement'' means to me.} that are
\textit{100\% semantically equivalent}: that is, given identical random seeds, a generator written with our technique
produces exactly the same sequence of values as the same generator written with the original generator library.
The semantic equivalence between the original generators and \name{} generators allows us to
perform clean, fine-grained comparisons, as differences in generation speed are
solely attributable to the staging and \rand optimizations.
\jwc{BCP thinks this should be a little stronger: ``and not obscured by the probablistic nature of generation'', or something like that. ``Comparing distributions of speed is hard, comparing pointwise is easier.''. This needs to have its own sentence here.}

>>>>

This is due to two major sources of inefficiency.
First, the same high-level design approach that gives generator libraries their
flexibility also introduces layered abstractions and indirections that are difficult for
compilers to optimize, shifting work to run time.
Second, the \rand that a generator uses to make choices can
constitute an unexpectedly high proportion of its runtime. These factors combine to limit 
the effectiveness of PBT.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi

We lay a path forward.
First, we show how to use \textit{multi-stage programming}---a classic technique for 
building fast DSLs “without regret”~\cn{}---to eliminate the overhead of 
many common generator abstractions.
Second, we show that entropy source is an undervalued consideration in generator library design
by isolating the performance contribution of the \rand in generation time, and providing an alternative
that behaves equivalently with $5.5\times$ speedups.
These insights culminate in \name, a recipe for building generator libraries that marry high-level design
with low-level performance.

% \bcp{It
% feels a bit grand to call what we've done an ``approach''---it's
% really just a single performance-enhancing technique, staging, plus
% the observation that the RNG matters and you should use the fastest
% one you can.}

To evaluate our technique, we implement staged PBT libraries in two strict functional languages: OCaml and Scala 3.
In OCaml, we build \camlname{}, based on \bq{}---a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
In Scala 3, we build \scalaname{}, based on ScalaCheck---a very
mature PBT library and the {\em de facto} choice\bcp{the wording here
  implies that there are other choices in Scala, some of which might
  be reasonable, and tht base\_quickcheck is not the de facto choice
  in OCaml...} in Scala \cn{}.
In both languages, our libraries are direct drop-in
replacements\bcp{that's not quite accurate, is it? I mean, they have
  the same functionality, but can you take a generator that works with
base\_quickcheck and just link it against our version without changing
the generator itself? That's what ``drop-in replacement'' means to me.} that are
\textit{100\% semantically equivalent}: that is, given identical random seeds, a generator written with our technique
produces exactly the same sequence of values as the same generator written with the original generator library.
The semantic equivalence between the original generators and \name{} generators allows us to
perform clean, fine-grained comparisons, as differences in generation speed are
solely attributable to the staging and \rand optimizations.
\jwc{BCP thinks this should be a little stronger: ``and not obscured by the probablistic nature of generation'', or something like that. ``Comparing distributions of speed is hard, comparing pointwise is easier.''. This needs to have its own sentence here.}
